Example of using mitch package for Infinium methylation analysis

### 2024-01-15 example_workflow/example_workflow.Rmd
instructions
refining the example workflow

---
title: "Example of using mitch package for Infinium methylation analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

CORES=2

@@

The limma data table is read in.
This dataset is available from GEO (GSE49667).
The limma data was obtained using the "A cross-package Bioconductor workflow for analysing methylation array data" workflow.

@@

```
## Load pathways

Reactome pathways were downloaded on the 14th Sept 2023 from MsigDB.

@@

Use all probes on the HM450K chip.

Updating gene names using the HGNC package because a lot of them have been updated over time.

An example of making a gene table for EPIC data is included down below.

@@

# mitch aggregate function needs to be changes to `mean`
tic()
m <- mitch_import(x=dm,DEtype="limma",geneTable=gt1,geneIDcol="Name")
toc() #2.5 sec

@@

Note that the results are not sorted by p-value, rather S.distance, an enrichment score.

@@

tic()
mres <- mitch_calc(x=m,genesets=gene_sets,minsetsize=5,cores=4, priority="effect")
toc() #7 sec on 4 threads

@@

@@

barnames <- gsub("_"," ",gsub("REACTOME_","",top$set))
cols <- as.character(((sign(top$s.dist)+1)/2)+1)
cols <- gsub("1","blue",gsub("2","red",cols))

par(mar = c(5.1, 29.1, 4.1, 2.1))
barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
mtext("whole gene")
grid()

@@

@@

@@

>>>>

---
title: "Example of using mitch package for Infinium methylation analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
bibliography: references.bib
csl: epigenetics.csl
---

@@

@@

The limma[@Ritchie2015-zn] data table is read in.
This dataset is available from GEO (GSE49667)[@Zhang2013-oy].
The limma data was obtained using a previously prepared workflow [@Maksimovic2017-zv].

@@

Note that this object has probes with their own column, not simply as row names.

## Load pathways

Gene ontologies were downloaded in GMT format from MSigDB on 15th Jan 2024[@Liberzon2015-um;@The_Gene_Ontology_Consortium2023-fm].
The GMT file is read into R using the mitch function `gmt_import()`.

@@

One of the critical parts of this workflow is the establishment of probe-gene relationships.
This controls how the probe data is aggregated to make the gene level scores.

In this demonstration we are using all probes on the HM450K chip.
Alternative approaches could use only promoters CpGs for example.

As these annotations are several years old, many of the annotated gene names are no longer
current.
To remedy this, the gene names are screened with the HGNChelper package and any defunct
symbols get updated to the newer gene name, so they will be recognised properly in the
gene sets.

The process for making a gene table for EPIC data is included down below.

@@

tic()
m <- mitch_import(x=dm,DEtype="limma",geneTable=gt1,geneIDcol="Name")
toc() #2.5 sec

@@

Note that the results are not sorted by p-value, rather S.distance, an enrichment score.
I think this works better for interpretation.

@@

tic()
mres <- mitch_calc(x=m,genesets=gene_sets,minsetsize=5, priority="effect")
toc() #7 secs

@@

Now make a barplot of these top findings.

```{r, barplot}

@@

barnames <- gsub("_"," ",top$set)
cols <- as.character(((sign(top$s.dist)+1)/2)+1)
cols <- gsub("1","blue",gsub("2","red",cols))

par(mar = c(5.1, 29.1, 4.1, 2.1))
barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Gene Ontologies")
mtext("whole gene")
grid()

@@

There are some interesting results there that give hints to the differences between naive T and Treg cells.

@@

In case you are working with EPIC array data, the probe-gene table can be make with this snippet of code.

@@

## Bibliography
